---
title: "Virat Kohli Test Career Analysis"
author: "Deepak Kumar"
date: "2025-05-13"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Load neccessary libraries
```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)

```

#level-1 data cleaning and preprocessing 

#load the dataset
```{r}
data<-read_excel("C:/virat kohli analysis/testcareermatchwise.xlsx")
View(data)
```


# Rename columns for readability
```{r}
colnames(data) <- c("Bat1", "Bat2", "Runs", "BallsFaced", "StrikeRate", "Fours", "Sixes", 
                    "Unused", "Opposition", "Ground", "StartDate", "TestNumber")

```




# Drop unused column
```{r}
data <- data %>% select(-Unused)

```




# Remove "v " prefix from Opposition
```{r}
data$Opposition <- gsub("^v\\s+", "", data$Opposition)

```



# Convert StartDate to Date format
```{r}
library(lubridate)

data$StartDate <- dmy(data$StartDate)


```




# Handle missing values and extract not out flag
```{r}
clean_innings <- function(score) {
  score <- as.character(score)
  score[score %in% c("DNB", "TDNB", "absent")] <- NA
  score <- gsub("\\*", "", score)  # Remove not-out asterisk
  return(as.numeric(score))
}
not_out_flag <- function(score) {
  score <- as.character(score)
  return(grepl("\\*", score))
}

```




# Apply cleaning to Bat1 and Bat2
```{r}
data <- data %>%
  mutate(
    Bat1_NotOut = not_out_flag(Bat1),
    Bat2_NotOut = not_out_flag(Bat2),
    Bat1 = clean_innings(Bat1),
    Bat2 = clean_innings(Bat2)
  )

# Combine total runs in both innings
data <- data %>%
  mutate(
    TotalRuns = rowSums(cbind(Bat1, Bat2), na.rm = TRUE),
    TotalNotOuts = Bat1_NotOut + Bat2_NotOut,
    PlayedInnings = (!is.na(Bat1)) + (!is.na(Bat2)),
    Year = year(StartDate)
  )

```




# Clean other numeric columns
```{r}
data <- data %>%
  mutate(
    Runs = as.numeric(Runs),
    BallsFaced = as.numeric(BallsFaced),
    StrikeRate = as.numeric(StrikeRate),
    Fours = as.numeric(Fours),
    Sixes = as.numeric(Sixes)
  )


```



 Add placeholder for match duration 
```{r}
data$MatchDuration <- 5  # By default, Test match = 5 days

```





# Preview the cleaned data
```{r}
str(data)
head(data)

```
*after performing all cleanig check the data type of columns and retrieve first 8 data using head  function*


###Level -2 stastical analysis of my dataset
# 1. Overall career batting average (runs per innings)
```{r}
career_average <- sum(data$TotalRuns, na.rm = TRUE) / sum(data$PlayedInnings, na.rm = TRUE)

```
*this average may differ from real average because in real average for cricket matches consider not out inning then calculate*




# 2. Career not-out percentage
```{r}
not_out_percentage <- mean(data$TotalNotOuts > 0) * 100

```

# 3. Highest score in a single innings
```{r}
highest_score <- max(c(data$Bat1, data$Bat2), na.rm = TRUE)

```
*this score was made by kohli against south africa* 



# 4. Total career runs
```{r}
total_runs <- sum(data$TotalRuns, na.rm = TRUE)

```
*in test matches virat has scored total 9230 runs*



# 5. Total centuries (100+ in any single innings)
```{r}
centuries <- sum(data$Bat1 >= 100, na.rm = TRUE) + sum(data$Bat2 >= 100, na.rm = TRUE)

```
*virat has scroed total 30 test centuries*



# 6. Total fifties (50–99 in any innings)
```{r}
fifties <- sum(data$Bat1 >= 50 & data$Bat1 < 100, na.rm = TRUE) +
           sum(data$Bat2 >= 50 & data$Bat2 < 100, na.rm = TRUE)

```
*virat has scored 31 total fifites*




# 7. Average runs per year
```{r}
avg_runs_year <- data %>%
  group_by(Year) %>%
  summarise(RunsInYear = sum(TotalRuns, na.rm = TRUE)) %>%
  mutate(AveragePerYear = round(RunsInYear / n(), 2))

```





# 8. Average against each opposition
```{r}
avg_opposition <- data %>%
  group_by(Opposition) %>%
  summarise(Average = round(sum(TotalRuns, na.rm = TRUE) / sum(PlayedInnings, na.rm = TRUE), 2),
            Matches = n())

```

*this table shows batting average of virat kohli against each teams*


# 9. Ground-wise batting average
```{r}
avg_ground <- data %>%
  group_by(Ground) %>%
  summarise(Average = round(sum(TotalRuns, na.rm = TRUE) / sum(PlayedInnings, na.rm = TRUE), 2),
            Matches = n())

```
*this table shows average of different grounds*




# 10. Strike Rate analysis: mean and max
```{r}
mean_strike_rate <- mean(data$StrikeRate, na.rm = TRUE)
max_strike_rate <- max(data$StrikeRate, na.rm = TRUE)
print(mean_strike_rate)
print(max_strike_rate)
```
*virat mean strike rate is 51.20223*
*virat max strike rate in test is 105.55*


###Level-3 visualization of my dataset
#loading required libraries visualization
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

```



#virat kohli runs against opponents 
```{r}
ggplot(data %>% group_by(Opposition) %>% summarise(TotalRuns = sum(TotalRuns, na.rm = TRUE)),
       aes(x = reorder(Opposition, -TotalRuns), y = TotalRuns)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(title = "Total Runs vs Each Opposition", x = "Opposition", y = "Runs")

```



#vira kohli batting average per oppositon 
```{r}
data %>%
  group_by(Opposition) %>%
  summarise(Average = sum(TotalRuns, na.rm = TRUE) / sum(PlayedInnings, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(Opposition, -Average), y = Average)) +
  geom_col(fill = "purple") +
  coord_flip() +
  labs(title = "Batting Average vs Opposition", x = "Opposition", y = "Average")

```


# virat kohli performence over time innings-by-innings 
```{r}
data %>%
  mutate(MatchNumber = row_number()) %>%
  ggplot(aes(x = MatchNumber, y = TotalRuns)) +
  geom_line(color = "blue") +
  geom_smooth(se = FALSE, color = "red") +
  labs(title = "Match-by-Match Total Runs", x = "Match Number", y = "Runs")

```



# virat kohli runs distributions  using histogram
```{r}
ggplot(data, aes(x = TotalRuns)) +
  geom_histogram(binwidth = 10, fill = "orange", color = "black") +
  labs(title = "Runs Distribution per Match", x = "Runs", y = "Frequency")

```


# virat kohli strike rate over time 
```{r}
data %>%
  mutate(MatchNumber = row_number()) %>%
  ggplot(aes(x = MatchNumber, y = StrikeRate)) +
  geom_line(color = "darkred") +
  geom_point() +
  labs(title = "Strike Rate Match-by-Match", x = "Match Number", y = "Strike Rate")

```


#virat kohli runs by ground  
```{r}
data %>%
  group_by(Ground) %>%
  summarise(Runs = sum(TotalRuns, na.rm = TRUE)) %>%
  arrange(desc(Runs)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(Ground, -Runs), y = Runs)) +
  geom_col(fill = "navy") +
  coord_flip() +
  labs(title = "Top 10 Grounds with Most Runs", x = "Ground", y = "Runs")

```



# virat kohli total fours and sixes
```{r}
data %>%
  summarise(TotalFours = sum(Fours, na.rm = TRUE),
            TotalSixes = sum(Sixes, na.rm = TRUE)) %>%
  pivot_longer(cols = everything(), names_to = "Type", values_to = "Count") %>%
  ggplot(aes(x = Type, y = Count, fill = Type)) +
  geom_col() +
  labs(title = "Total Fours vs Sixes", x = "Type", y = "Count")

```


#virat kohli  bat1 vs bat2 comparison  using  bar plot 
```{r}

data <- data %>%
  mutate(Match = row_number())

data_long <- data %>%
  select(Match, Bat1, Bat2) %>%
  pivot_longer(cols = c(Bat1, Bat2), names_to = "Innings", values_to = "Runs")

data_long$Runs <- as.numeric(data_long$Runs)

ggplot(data_long, aes(x = Match, y = Runs, color = Innings)) +
  geom_line(size = 1) +
  labs(title = "Innings-wise Batting Comparison", x = "Match Number", y = "Runs Scored") +
  theme_minimal()


```


# 100s and 50s highlight chart
```{r}
ggplot(data, aes(x = Match, y = TotalRuns, fill = TotalRuns)) +
  geom_col() +
  scale_fill_gradient2(low = "skyblue", high = "darkred", midpoint = 100) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "orange") +
  geom_hline(yintercept = 100, linetype = "dashed", color = "green") +
  labs(title = "Test Scores with 50 & 100 Milestones", x = "Match", y = "Total Runs") +
  theme_minimal()

```


##Runs Over Time
```{r}


data <- data %>%
  mutate(CumulativeRuns = cumsum(TotalRuns))

ggplot(data, aes(x = Match, y = CumulativeRuns)) +
  geom_line(color = "darkgreen", size = 1) +
  labs(title = "Cumulative Runs Over Test Career", x = "Match Number", y = "Cumulative Runs") +
  theme_minimal()

```
*this graph shows total runs have accumulayed across his career*
